The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-11 rational.tex
done

\paragraph*{Lifting negation out of {{*}}.}
The isomorphisms below state that the direction is \emph{relative}. If {{b1}}
and {{b2}} are flowing opposite to each other then it doesn't matter which is
direction is forwards and which is backwards.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Future work and Conclusion}
\label{sec:conc}

@@

In another sense however, this paper is about the nature of duality in
computation. The concept of duality is deep and significant and in
many cases -- probably most famously in the divide between classical
and intuinionistic mathematics -- contain non-trivial assumptions
about our world-view.  The problem of duality in computation and
logic, has many facets of which `continuation' and `higher order
functions' are only one.  We have contributed an elegant and natural
notion of duality that arises from the monoidal structure of sums and
pairs. This notion of duality has a crisp semantics, clear
computational interpretations and an information theoretic basis.
Before we conclude, we briefly mention some closely related ideas that
warrant further study.

@@

\paragraph*{Geometry of Interaction.}
GoI was developed by Girard \cite{girard1989geometry}as basis for the
proof theory of Linear Logic.  GoI however has been developed into a
reversible model of computing
\cite{Mackie2011,DBLP:conf/popl/Mackie95,Abramsky:1994:NFG:184662.184664}
in which values traveling along wires, much like {{langRevEE}}.
Preliminary investigations suggest that many GoI machine constructions
can be simulated in {{langRevEE}} by treating Mackie's bi-directional
wires as pairs of wires in {{langRevEE}} and replacing the machine's
global state with a typed value on the wire that captures the
appropriate state.  However a concrete connection between GoI machines
and GoI in general requires further work. This connection is exciting
because when viewed through a Curry-Howard lens it suggests that the
logical interpretation of {{langRevEE}} should a logic much like LL
with a strong notion of resource preservation, but also with a strong
computational basis.

\paragraph*{Computing in the Field of Algebraic Numbers.}
\label{sec:algebraic-field}
The field of algebraic numbers consist of the rationals extended with
solutions to polynomials. In a limited sense {{langRevEE}} extended
with recursive types already include polynomials such as the
definition of binary tree {{@@x.(1+x*x)}} implies the equality
{{x=1+x*x}} and hence {{x^2-x+1=0}}.

The famous example of isomorphism on algebraics is the embedding of
seven binary trees in one. In Figure 1 of Fiore's paper
\cite{Fiore:2004}, he demonstrates this isomorphism by the
corrsponding algebraic manipulation of the types.  The algebraic
manipulation can be viewed as a {{langRevEE}} combinator and it indeed
does constructively witness the isomorphism, i.e., you can run it and
view one data structure encoded by the other. In this context,
consider the following algebraically valid proof of the same
isomorphism:
\[\begin{array}{rclclclcl}
x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
x^6 &=& 1 \\
x^7 &=& x^6 x &=& x
\end{array}\]
Fiore poses the question of why such algebraic manipulation would make
sense type theoretically, even though some of the intermediate steps
make no sense.  Stated in terms of {{langRevEE}} computations, this
manipulation has no meaningful computational content -- it is always
non-terminating. It suggests that some algebraic manipulations are
somehow more `constructive' than others or dually that {{langRevEE}}
with recursive types somehow lacks the ability to express all
computations that are otherwise algebraically meaningful.

\begin{itemize}
\item Maybe the problem is that there is cancellation in the second
  case and cancellation with recursive types is problematic. However
  not all cancellation is. So what gives?

\item One idea is that irrationals correspond to an infinite amount of
  computational work. If what we are cancelling is not the same
  infinity, then things go wrong.

\item Chaitin has already shown that reals have inifite amount of
  information which is paradoxical and problematic. This is a
  different issue from that of the rational being infinite
  computations.

\item Finally, not all meaningful types are meaningful polynomials --
  ex. nat. So this is not the whole story.

\paragraph*{Quantum Computing} 
A \emph{time traveling} intuition is also applicable here.  In a
normal cicruit, as computational steps are taken values flow from left
to right, i.e. computational time progresses from left to right.  The
action of {{eps+}} causes values, aka information particles, to flow
from right to left i.e. backwards in computational time.  The
interesting thing about this interpretation is that particles that
travel backwards in time get to see and interact with the history of
particles that they coexist with. This gives us an intuitive
interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
Sec. \ref{sec:specific-constructions}) where it would otherwise seem
that {{-b1}} and {{b2}} move in opposite directions. The backward flow
of the {{-b1}} allows it to `see the past' of {{b2}} and is thus
equivalent to both particles moving backward in time.

An analogy inspired by quantum mechanics is also applicable here.  The
operator {{eta*}} is a site of fission that creates \emph{an entangled
  particle and anti-particle}. These particles can be thought to be in
a \emph{superposition} of states determined by their type. Both flow
in the same direction in time.  In each possible world that the
particle exists, it takes on a specific value inhibiting its type and
correspondingly its entangled anti-particle takes the dual of the
specific value. Since the particles are entangled, actions on one
(such as transformation by application of an isomorphism) affects the
other in much the same sense as \emph{action at a distance}. The
`{{a-o*b}}' functions mentioned in the introduction, are first-class
values and they correspond to these entagled pairs. Finally {{eps*}}
operations are sites where corresponding particle and anti-particles
annihilate each other. While appealing as an analogy, this description
does not imply (or preclude) any formal connection with Physics.

Are constructions are very similar to those developed in the context
of QC. Is this how nature computes? Are the analogies to time
traveling particles and anti-particles more concrete in some way?

@@

%% Jacques Carette, Michael Adams, Khaled, Will Byrd, Yin Wang

\acks This project was partially funded by Indiana University's Office
of the Vice President for Research and the Office of the Vice Provost
for Research through its Faculty Research Support Program.  We also
acknowledge support from Indiana University's Institute for Advanced
Study.

>>>>

\paragraph*{Lifting negation out of {{*}}.}
The isomorphisms below state that the direction is \emph{relative}. If {{b1}}
and {{b2}} are flowing opposite to each other then it doesn't matter which is
direction is forwards and which is backwards. More interestingly as {{b1}} is
moving backwards, it can ``see the past'' of {{b2}} which is equivalent to
both particles moving backwards.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion and Future Work}
\label{sec:conc}

@@

In another sense however, this paper is about the nature of duality in
computation. 
% The concept of duality is deep and significant and in many cases ---
% probably most famously in the divide between classical and intuitionistic
% mathematics --- contain non-trivial assumptions about our worldview.  The
% problem of duality in computation and logic, has many facets of which
% `continuation' and `higher order functions' are only one.
The concept of duality is deep and significant: we have opened the door for
us to consider, not one but two notions of duality. Surprisingly this makes
things significantly simpler. In particular, instead of conflating
considering pairs as dual to sums, the tradition in mathematics has long been
to consider fields with two notions of duality: one for sums and one for
pairs. This double notion of duality has a crisp semantics, clear
computational interpretation, and an information theoretic basis.

Our work has barely scratched the surface of an area of computing which has
been explored in depth before but without the combined reversible
information-preserving framework and the two notions of duality. The new
insights point to further new areas of investigation of which we mention
the three most significant ones (in our opinion).

@@

\paragraph*{Geometry of Interaction (GoI.}
This geometry was developed by Girard~\cite{girard1989geometry} as part of
the development of linear logic. It was given a computational interpretation
by Abramsky and Jagadeesan~\cite{Abramsky:1994:NFG:184662.184664}, and was
developed into a reversible model of computing by
Mackie~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. Preliminary investigations
suggest that many of the GoI machine constructions can be simulated in
{{langRevEE}} by treating Mackie's bi-directional wires as pairs of wires in
{{langRevEE}} and replacing the machine's global state with a typed value on
the wire that captures the appropriate state. This connection is exciting
because when viewed through a Curry-Howard lens it suggests that the logical
interpretation of {{langRevEE}} would be a linear-like logic with a notion of
resource preservation and with a natural computational interpretation.

\paragraph*{Computing in the Field of Algebraic Numbers.}
\label{sec:algebraic-field}
Algebraically, the move from {{langRev}} to {{langRevEE}} corresponds to a
move from a ring-like structure to a full field. Our language {{langRevEE}}
captures the structure of one particular field: that of the rational
numbers. As we have seen, computation in this field is quite expressive and
interesting and yet, it has two fundamental limitations. First it cannot
express any recursive type, and second it cannot express any datatype
definitions. We believe these to be two orthogonal extensions: recursive
types were considered in our previous paper~\cite{infeffects}; arbitrary
datatypes are however even more exciting that plain rationals as each
datatype definition can be viewed as a polynomial (see below) which
essentially means that we start computing in the field of algebraic numbers,
which includes square roots and imaginary numbers. As crazy as it might seem,
the type $\sqrt{2}$ and even the type $(1/2)+i(\sqrt{3}/2)$ ``make sense.''
In fact the latter type is the solution to the polynomial {{x^2-x+1=0}} which
if re-arranged looks like {{x=1+x*x}} and perhaps more familiarly as the
datatype of binary trees {{@@x.(1+x*x)}}. These types happen to have been
studied extensively following a paper by Blass~\cite{seventrees} which used
the above datatype of trees to infer an isomorphism between seven binary
trees and one!

We have confirmed that we can extended {{langRevEE}} with the datatype
declaration for binary trees and build a witness for this isomorphism that
works as expected. To understand the issue in more detail, consider the
following algebraically valid proof of the isomorphism in question:
\[\begin{array}{rclclclcl}
x^3 &=& x^2 x &=& (x-1) x &=& x^2 - x &=& -1 \\
x^6 &=& 1 \\
x^7 &=& x^6 x &=& x
\end{array}\]
The question is why such an algebraic manipulation makes sense type
theoretically, even though the intermediate step asking for an isomorphism
between $x^6$ and $1$ has no computational context. In the setting of
{{langRevEE}}, this isomorphism can be constructed but it diverges on all
inputs (in both ways). This suggests that, in the field of algebraic numbers,
some algebraic manipulations are somehow 
``more constructive'' than others.
%% or dually that {{langRevEE}} with recursive types
%%somehow lacks the ability to express all computations that are otherwise
%%algebraically meaningful.

% \begin{itemize}
% \item Maybe the problem is that there is cancellation in the second
%   case and cancellation with recursive types is problematic. However
%   not all cancellation is. So what gives?

% \item One idea is that irrationals correspond to an infinite amount of
%   computational work. If what we are cancelling is not the same
%   infinity, then things go wrong.

% \item Chaitin has already shown that reals have inifite amount of
%   information which is paradoxical and problematic. This is a
%   different issue from that of the rational being infinite
%   computations.

% \item Finally, not all meaningful types are meaningful polynomials --
%   ex. nat. So this is not the whole story.

\paragraph*{Quantum Computing.} 
One understand of quantum computing is that it exploits the laws of physics
to build faster machines (perhaps). Another more foundational understanding
is that it provides a computational interpretation of physics, and in
particular directly address the question of interpretation of quantum
mechanics. In a little known document, Rozas~\cite{Rozas:1987:CMO:889539}
uses continuations to implement the transactional interpretation of quantum
mechanics~\cite{transactional} which includes as its main ingredient a
fixpoint calculation between waves or particles traveling forwards and
backwards in time. Our work shed no light on whether this interpretation is
the ``right one'' but it is interesting that we can directly realize it using
the primitives of {{langRevEE}}.

% A \emph{time
%   traveling} intuition is also applicable here.  In a normal cicruit, as
% computational steps are taken values flow from left to right,
% i.e. computational time progresses from left to right.  The action of
% {{eps+}} causes values, aka information particles, to flow from right to left
% i.e. backwards in computational time.  The interesting thing about this
% interpretation is that particles that travel backwards in time get to see and
% interact with the history of particles that they coexist with. This gives us
% an intuitive interpretation of the isomorphism {{(-b1)*b2<->-(b1*b2)}} (see
% Sec. \ref{sec:specific-constructions}) where it would otherwise seem that
% {{-b1}} and {{b2}} move in opposite directions. The backward flow of the
% {{-b1}} allows it to `see the past' of {{b2}} and is thus equivalent to both
% particles moving backward in time.

The multiplicative structure of {{langRevEE}} also has a direct connection to
entangled quantum particles, or perhaps entangled particles and
anti-particles.  The idea of entanglement is one action on one particle is
``instanteneously'' communicated to the other. Again our model shed no light
on whether this is related to how nature computes but it is again interesting
that we can directly realize the idea using the primitives of {{langRevEE}}.

% natural An analogy
% inspired by quantum mechanics is also applicable here.  The operator {{eta*}}
% is a site of fission that creates \emph{an entangled particle and
%   anti-particle}. These particles can be thought to be in a
% \emph{superposition} of states determined by their type. Both flow in the
% same direction in time.  In each possible world that the particle exists, it
% takes on a specific value inhibiting its type and correspondingly its
% entangled anti-particle takes the dual of the specific value. Since the
% particles are entangled, actions on one (such as transformation by
% application of an isomorphism) affects the other in much the same sense as
% \emph{action at a distance}. The `{{a-o*b}}' functions mentioned in the
% introduction, are first-class values and they correspond to these entagled
% pairs. Finally {{eps*}} operations are sites where corresponding particle and
% anti-particles annihilate each other. While appealing as an analogy, this
% description does not imply (or preclude) any formal connection with Physics.

% Are constructions are very similar to those developed in the context
% of QC. Is this how nature computes? Are the analogies to time
% traveling particles and anti-particles more concrete in some way?

@@

\acks We thank Jacques Carette for stimulating discussion, and Michael Adams,
Will Byrd, Lindsey Kuper, and Yin Wang for helpful comments and
questions. This project was partially funded by Indiana University's Office
of the Vice President for Research and the Office of the Vice Provost for
Research through its Faculty Research Support Program.  We also acknowledge
support from Indiana University's Institute for Advanced Study.
